---
title: 'Hypothesis H2 (P2): Communication Frequency impacts Communication Flaws'
author: "Julian Frattini"
date: "2025-12-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse) # R dialect for simpler data wrangling
library(patchwork) # arranging and assembling figures
library(ggdag) # visualizing directed acyclic graphs

library(brms) # training of Bayesian regression models
library(marginaleffects) # evaluation of Bayesian regression models

source("../util/data-loader.R") # preparation of the raw data
```

This document investigates hypothesis $H_2^{P2}$ (originating from proposition 2):

> *Communication frequency* cause *communication flaws within our project team*.

The two variables are concepts operationalized by the following self-reported variables:

> *The self-reported frequency of meeting team members* impacts whether there are *communication flaws within a project team*.

For analysis, we follow the 3-step framework for statistical causal inference as summarized by Siebert.[^1]

## Modeling

The causal relationships were formalized in the following directed, acyclic graph (see proposition 2 for the full DAG):

```{r specify-dag}
dag <- dagify(
  comflaw.team ~ frequency,

  exposure = "frequency",
  outcome = "comflaw.team",
  
  labels = c(
    frequency = "frequency", comflaw.team = "comflaw-team"
  ),
  coords = list(
    x = c(frequency = 0, comflaw.team = 1),
    y = c(frequency = 3, comflaw.team = 1.5)
  )
)

 ggdag_status(dag, use_labels = "label", text = FALSE) +
   theme_dag() +
   theme(legend.position = "bottom") +
   labs(fill = "Variable type", color = "Variable type")
```

## Identification

Based on the backdoor criterion, we can identify variables to statistically control

```{r identify-adjustment-set}
ggdag_adjustment_set(dag, exposure = "frequency", outcome = "comflaw.team")
```

None of the variables considered in the DAG confound the relationship of interest.
This yields the following set of relevant variables:

| Code | Group | Description | Variable |
|---|---|---|---|
| `frequency` | Communication | How often do members of your team meet, informally by chance meeting or in announced meetings? | `v_380` |
| `comflaw.team` | Problems | Communication flaws within our project team | `v_385` |

## Estimation

In the third step, we quantify the strength of the assumed causal effect.
We employ a Bayesian data analysis approach following the guidelines of McElreath[^2] and Gelman et al.[^3]

### Data Preparation

We load the data from the 2024 run of the NaPiRE survey.
The variables relevant to this hypothesis are LIST.
We use the file *data_cleaned_2024.csv*, not the original *data_raw_2024.csv*, because `v_3` is a numeric variable but the original, raw file allowed strings (e.g., "12 people" instead of $12$) in this field.
In the process, we make the following changes:

- Dropping incomplete data: removing all observations containing `NA` values
- Dropping "I don't know": rows where `v_380` (i.e., the self-reported communication frequency) is "I don't know" provide no information to our analysis

All manual cleaning operations performed on the data are documented in the data cleaning protocol (to be found in data/runs/napire2024/data-cleaning-protocol.md).

```{r load-data}
d <- load_data(run = "2024", file = "../../data/runs/napire2024/data_cleaned_2024.csv") %>% 
  select(v_380, v_385) %>%  
  drop_na() %>% 
  filter(v_380 != "I don't know")

str(d)
```

Before starting with the analysis, we can visualize the distribution of the data.
Ordinal variables are visualized as stacked bars.

```{r visualize-data}
plot.distribution.v380 <- ggplot(d, aes(x = v_380, fill = v_380)) +
    geom_bar() +
    labs(x = "Meeting frequency", y = "Count") +
    theme_bw() +
    guides(fill = "none")

plot.distribution.v385 <- ggplot(d, aes(x = v_385, fill = v_385)) +
    geom_bar() +
    labs(x = "Perceiving that there are communication flaws in the team", y = "Count") +
    theme_bw() +
    guides(fill = "none")

plot.distribution.v380 / plot.distribution.v385
```

### Formula

Next, we specify a regression formula based on the identification step.
The outcome variable is ordinal (since it is measured on a 5-point Likert-scale).
We can assume that the ordinal variable is an ordinal operationalization of an unmeasurable latent variables on a continuous scale, i.e., the true agreement to the statement that changes in requirements are not communicated.
Hence, we use the `brms::cumulative` distribution family with a Probit link function.

```{r define-formula}
f1 <- bf(v_385 ~ 1 + v_380)

get_prior(f1, data = d, family = cumulative("probit"))
```

### Priors

We select uninformative priors for all coefficients, i.e., 0-centered normal distributions with wide variance.

```{r set-priors}
p1 <- c(
  set_prior("normal(0, 1)", class = "Intercept"), 
  set_prior("normal(0, 0.5)", class = "b")
)
```

To assess the eligibility of these priors, we perform a prior predictive check.
During this check, we sample from the model using the prior probability distributions for each factor coefficient without updating it in light of the data.

```{r prior-predictive-check}
m1.prior <-
  brm(data = d, family = cumulative("probit"), formula = f1, prior = p1,
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    seed = 4, sample_prior="only",
    file = "../fits/p2h2.prior", file_refit = "on_change"
  )

brms::pp_check(m1.prior, ndraws=100, type = "bars")
```

The sampled prior predictive checks (black whiskers labeled $y_{rep}$) overlap the actually observed data (blue bars labeled $y$), indicating that the selected priors entail the observed reality as realistic.

### Training

With the eligibility of the priors confirmed, we can train the Bayesian regression model, i.e., this time updating the coefficient distributions using Bayes' theorem.

```{r model-training}
m1 <-
  brm(data = d, family = cumulative("probit"), formula = f1, prior = p1,
    iter = 2000, warmup = 1000, chains = 4, cores = 4, seed = 4, 
    file = "../fits/p2h2", file_refit = "on_change"
  )
```

The training does not mention any diverging chains, which indicates solid sampling.
To assess that the model has learned, we perform a posterior predictive check.

```{r posterior-predictive-check}
brms::pp_check(m1, ndraws=100, type="bars")
```

The posterior predictions $y_rep$ still overlap the actually observed data $y$ but the whiskers have grown closer to the data, which indicates that the model has learned.

### Evaluation

Finally, we can evaluate the results to answer the initially posed hypothesis.

#### Posterior Distributions

First, we can investigate the raw posterior distributions obtained from the model training.
These distributions represent the effect that the respective factor has on the outcome variable.
If a distribution is not consistent with 0 (i.e., it does not overlap with 0), then it can be considered statistically significant.

```{r print-posteriors}
summary(m1)
```

These distributions are, however, difficult to interpret, because they are in the parameter space (i.e., on a probit scale), not the outcome space.
To investigate the effects on the outcome space (i.e., in terms of how they affect the likelihood to select a certain category of the Likert-scale), we need to assess the marginal effects.

#### Marginal Effects

To visualize the effect of the predictors on the outcome variable, we plot the marginal effects.
Marginal effects isolate the effect of one predictor while keeping all other predictors at a representative level.
The main effect of interest is whether communication frequency (`v_380`) impacts the agreement to the statement that a survey participant perceives communication flaws within their team (`v_385`).
In the resulting plot, each dot and overlapping whisker represents the mean value and the credibility interval for the probability (y-axis) that a survey participant that meets their team with a given frequency (x-axis) would give a certain response (color) to the question "How much do you agree or disagree with the following statements about your project: Communication flaws within our project team." (`v_385`).

```{r marginal-frequency}
conditional_effects(m1, "v_380", categorical = TRUE, prob = 0.89)[[1]] %>%
  ggplot(aes(x = v_380, color = cats__)) +
    geom_point(aes(y = estimate__), size = 2, position = position_dodge(width = 0.5)) +
    geom_errorbar(aes(ymin = lower__, ymax = upper__), width = 0.3, position = position_dodge(width = 0.5)) +
    labs(x = "Frequency of meeting the team", y = "Probability", color = "Response") +
    scale_y_continuous(labels = scales::percent) +
    theme_bw() +
    theme(legend.position = "bottom")
```

The marginal effect shows that when the frequency of meeting the team increases, the probability of agreeing to the statement increases while the probability of disagreeing with the statement decreases.
Visualizing only the average probability (without the credibility interval), dropping the "No opinion"-level, and separating the two agreeing categories from the two disagreeing categories produces the following plot that makes this point even more clear.

```{r marginal-frequency-simplified}
conditional_effects(m1, "v_380", categorical = TRUE, prob = 0.89)[[1]] %>% 
  filter(cats__ != "No opinion") %>% 
  mutate(agree = ifelse(cats__ %in% c("Strongly Agree", "Agree"), "Agreeing", "Disagreeing")) %>% 
  ggplot(aes(x = v_380, color = cats__)) +
    geom_point(aes(y = estimate__), size = 2) +
    geom_line(aes(y = estimate__, group = cats__), alpha = 0.7) +
    labs(x = "Frequency of meeting the team", y = "Average probability", color = "Response") +
    scale_y_continuous(labels = scales::percent) +
    facet_wrap(~agree, ncol = 1) +
    theme_bw()
```

Note that this plot connects the dots (i.e., the average probability per frequency) with lines, which is mathematically incorrect, as the x-axis is ordinal (i.e., categorical) and there is no continuum between two levels.
The connecting lines are solely to visualize the trend better.

## Conclusion

The obtained results invite the following conclusion: 
The more frequently software engineering teams meet, the fewer communication flaws they perceive.
This is consistent with the intuition that regular and frequent exchange of information mitigates communication problems.

[^1]: Siebert, J. (2023). Applications of statistical causal inference in software engineering. Information and Software Technology, 159, 107198.
[^2]: McElreath, R. (2018). Statistical rethinking: A Bayesian course with examples in R and Stan. Chapman and Hall/CRC.
[^3]: Gelman, A., Carlin, J. B., Stern, H. S., & Rubin, D. B. (1995). Bayesian data analysis. Chapman and Hall/CRC.