---
title: 'Hypothesis H1b (P2): Communication Frequency impacts Change Communication'
author: "Julian Frattini, Stefan Wagner"
date: "2025-12-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse) # R dialect for simpler data wrangling
library(patchwork) # arranging and assembling figures
library(ggdag) # visualizing directed acyclic graphs

library(brms) # training of Bayesian regression models
library(marginaleffects) # evaluation of Bayesian regression models

source("../util/data-loader.R") # preparation of the raw data
```

This document investigates hypothesis $H_{1b}^{P2}$ (originating from proposition 2):

> The *way-of-working* moderates how *communication frequency* impacts that *changes in requirements are not communicated* .

This hypothesis emerged from the results to hypothesis $H_1^{P2}$, which showed a non-monotonic relationship between communication frequency and the perception that changes are not communicated:
Both teams that meet very often (e.g., almost daily) and very rarely (e.g., bi-monthly) report less problems with change communication than teams meeting bi-weekly.
There are two potential revisions to the original causal assumption that might explain this result:

1. The observed effect is an overlapping effect from two distinct classes of software development projects separated by their *way-of-working*: for teams working in an agile way, meeting bi-weekly may be too infrequent, for teams working in a plan-driven manner, it may be too frequent.
2. Projects with less frequent meetings entail stricter *documentation*, which - in turn - improves communication of changes.

In this analysis of $H_{1b}^{P2}$, we revise the original causal assumptions to check whether the additional assumptions explain the observed non-monotonic effect in $H_1^{P2}$.

For analysis, we follow the 3-step framework for statistical causal inference as summarized by Siebert.[^1]

## Modeling

The causal relationships were formalized in the following directed, acyclic graph (see proposition 2 for the baseline DAG):

```{r specify-dag}
dag <- dagify(
  team.size ~ wow,
  frequency ~ team.size + wow,
  changes ~ frequency + team.size + wow + doc,
  doc ~ frequency,

  exposure = "frequency",
  outcome = "changes",
  latent = "doc",
  
  labels = c(
    frequency = "frequency", changes = "changes", team.size = "team-size", 
    wow = "way-of-working", doc = "documentation"
  ),
  coords = list(
    x = c(frequency = 0, changes = 0.5, team.size = 0, wow = 0.25, doc = 0.25),
    y = c(frequency = 3, changes = 1.5, team.size = 0.5, wow = -1, doc = 3)
  )
)

 ggdag_status(dag, use_labels = "label", text = FALSE) +
   theme_dag() +
   theme(legend.position = "bottom") +
   labs(fill = "Variable type", color = "Variable type")
```

The three relevant variables in this hypothesis are:

| Code | Group | Description | Variable |
|---|---|---|---|
| `frequency` | Communication | How often do members of your team meet, informally by chance meeting or in announced meetings? | `v_380` |
| `changes` | Problems | How much do you agree or disagree with the following statements about your project: Changes in requirements are not communicated | `v_410` |
| `team-size` | Demographics | How many people are involved in your project? | `v_3` |
| `way-of-working` | Demographics | How would you personally characterize the way your project works (agile vs. plan-driven)? | `v_457` |
| `documentation` | - | How excensively do you document the requirements in your project? | |

The variable `documentation` was not observed in the 2024 run of the survey.

## Identification

Based on the backdoor criterion, we can identify variables to statistically control

```{r identify-adjustment-set}
ggdag_adjustment_set(dag, exposure = "frequency", outcome = "changes")
```

We must adjust for the variable `team.size` and `way-of-working` to de-confound the effect of `frequency` on `changes`.

## Estimation

In the third step, we quantify the strength of the assumed causal effect.
We employ a Bayesian data analysis approach following the guidelines of McElreath[^2] and Gelman et al.[^3]

### Data Preparation

We load the data from the 2024 run of the NaPiRE survey.
The variables relevant to this hypothesis are `team.size` (`v_3`), `frequency` (`v_380`), and `changes` (`v_410`), and `wow` (`v_457`).
We use the file *data_cleaned_2024.csv*, not the original *data_raw_2024.csv*, because `v_3` is a numeric variable but the original, raw file allowed strings (e.g., "12 people" instead of $12$) in this field.
In the process, we make the following changes:

- Dropping incomplete data: removing all observations containing `NA` values
- Add log-teamsize: creating a separate variable $v\_3l = \log_{10}(v\_3)$ that reflects the *magnitude* of team size, not the raw team size itself
- Removing "don't know": removing rows where `v_380` is "I don't know", as this data does not have any value to our analysis.

All manual cleaning operations performed on the data are documented in the data cleaning protocol (to be found in data/runs/napire2024/data-cleaning-protocol.md).

```{r load-data}
d <- load_data(run = "2024", file = "../../data/runs/napire2024/data_cleaned_2024.csv") %>% 
  select(v_3, v_380, v_410, v_457) %>%  
  drop_na() %>% 
  mutate(v_3l = log(v_3, base = 10)) %>% 
  filter(v_380 != "I don't know") %>% 
  filter(v_457 != "I don't know")

str(d)
```

Before starting with the analysis, we can visualize the distribution of the data.
Ordinal variables are visualized as stacked bars.

```{r visualize-data-ordinal}
plot.distribution.v380 <- ggplot(d, aes(x = v_380, fill = v_380)) +
    geom_bar() +
    labs(x = "Meeting frequency", y = "Count") +
    theme_bw() +
    guides(fill = "none")

plot.distribution.v410 <- ggplot(d, aes(x = v_410, fill = v_410)) +
    geom_bar() +
    labs(x = "Requirements changes are not communicated", y = "Count") +
    theme_bw() +
    guides(fill = "none") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot.distribution.v457 <- ggplot(d, aes(x = v_457, fill = v_457)) +
    geom_bar() +
    labs(x = "Way of Working", y = "Count") +
    theme_bw() +
    guides(fill = "none") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot.distribution.v380 / (plot.distribution.v410 | plot.distribution.v457)
```

We can visualize the numeric data (i.e., `v_3` and its logarithmized form `v_3l`) with histograms.

```{r visualize-data-numeric}
plot.distribution.v3 <- ggplot(d, aes(x = v_3)) +
  geom_histogram(bins = 40) +
  labs(x = "Team size", y = "Count") +
  theme_bw()

plot.distribution.v3l <- ggplot(d, aes(x = v_3l)) +
  geom_histogram(bins = 40) +
  labs(x = "Magnitude of team size (log_10(team.size))", y = "Count") +
  theme_bw()

plot.distribution.v3 / plot.distribution.v3l
```

The logarithm of `team.size` $\log_{10}(v\_3)$ shows a Gaussian-looking distribution of team size magnitude.
Both distributions represent the same variable and same data, just on different scales.

### Formula

Next, we specify a regression formula based on the identification step, i.e., when regressing `v_410` on `v_380`, we statistically control for the confounders `v_3l` and `v_457`.
The outcome variable is ordinal (since it is measured on a 5-point Likert-scale).
We can assume that the ordinal `v_410` is an ordinal operationalization of an unmeasurable latent variables on a continuous scale, i.e., the true agreement to the statement that changes in requirements are not communicated.
Hence, we use the `brms::cumulative` distribution family with a Probit link function.
We assume that the *way-of-working* (`v_457`) moderates the effect of *meeting frequency* (`v_380`) on *communication of changes* (`v_410`), which entails an interaction effect `v_380*v_457`.

```{r define-formula}
f1 <- bf(v_410 ~ 1 + v_380*v_457 + v_3l)

# print the full list of available priors
# get_prior(f1, data = d, family = cumulative("probit"))
```

### Priors

We select uninformative priors for all coefficients, i.e., 0-centered normal distributions with wide variance.

```{r set-priors}
p1 <- c(
  set_prior("normal(0, 1)", class = "Intercept"), 
  set_prior("normal(0, 0.5)", class = "b")
)
```

To assess the eligibility of these priors, we perform a prior predictive check.
During this check, we sample from the model using the prior probability distributions for each factor coefficient without updating it in light of the data.

```{r prior-predictive-check}
m1.prior <-
  brm(data = d, family = cumulative("probit"), formula = f1, prior = p1,
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    seed = 4, sample_prior="only",
    file = "../fits/p2h1b.prior", file_refit = "on_change"
  )

brms::pp_check(m1.prior, ndraws=100, type = "bars")
```

The sampled prior predictive checks (black whiskers labeled $y_{rep}$) overlap the actually observed data (blue bars labeled $y$), indicating that the selected priors entail the observed reality as realistic.

### Training

With the eligibility of the priors confirmed, we can train the Bayesian regression model, i.e., this time updating the coefficient distributions using Bayes' theorem.

```{r model-training}
m1 <-
  brm(data = d, family = cumulative("probit"), formula = f1, prior = p1,
    iter = 2000, warmup = 1000, chains = 4, cores = 4, seed = 4, 
    file = "../fits/p2h1b", file_refit = "on_change"
  )
```

The training does not mention any diverging chains, which indicates solid sampling.
To assess that the model has learned, we perform a posterior predictive check.

```{r posterior-predictive-check}
brms::pp_check(m1, ndraws=100, type="bars")
```

The posterior predictions $y_rep$ still overlap the actually observed data $y$ but the whiskers have grown closer to the data, which indicates that the model has learned.

### Evaluation

Finally, we can evaluate the results to answer the initially posed hypothesis.

#### Posterior Distributions

First, we can investigate the raw posterior distributions obtained from the model training.
These distributions represent the effect that the respective factor has on the outcome variable.
If a distribution is not consistent with 0 (i.e., it does not overlap with 0), then it can be considered statistically significant.

```{r print-posteriors}
summary(m1)
```

These distributions are, however, difficult to interpret, because they are in the parameter space (i.e., on a probit scale), not the outcome space.
To investigate the effects on the outcome space (i.e., in terms of how they affect the likelihood to select a certain category of the Likert-scale), we need to assess the marginal effects.

#### Marginal Effects

To visualize the effect of the predictors on the outcome variable, we plot the marginal effects.
Marginal effects isolate the effect of one predictor while keeping all other predictors at a representative level.
First, we visualize the effect of *meeting frequency* on *communication problems*.
In the resulting plot, each dot and overlapping whisker represents the mean value and the credibility interval for the probability (y-axis) that a survey participant that meets their team with a given frequency (x-axis) would give a certain response (color) to the question "How much do you agree or disagree with the following statements about your project: Changes in requirements are not communicated." (`v_410`).

```{r marginal-frequency}
conditional_effects(m1, "v_380", categorical = TRUE, prob = 0.89)[[1]] %>% 
  ggplot(aes(x = v_380, color = factor(cats__, levels = cat.likert, ordered = TRUE))) +
    geom_point(aes(y = estimate__), size = 2, position = position_dodge(width = 0.5)) +
    geom_errorbar(aes(ymin = lower__, ymax = upper__), width = 0.3, position = position_dodge(width = 0.5)) +
    labs(x = "Frequency of meeting the team", y = "Probability", color = "Change not communicated") +
    scale_y_continuous(labels = scales::percent) +
    theme_bw() +
    theme(legend.position = "bottom")
```

The figure shows a similar effect as observed in $H_1^{P2}$, but with two differences:

1. The effect is more pronounced, now.
2. The sour spot moved from the meeting frequency `bi-weekly` to `weekly`.

A filtered plot just focusing on the average response probabilities (i.e., without the whisker plot), leaving out the "No opinion"-option, and separating the two agreeing from the two disagreeing levels makes this effect even more clear.

```{r marginal-frequency-simplified}
conditional_effects(m1, "v_380", categorical = TRUE, prob = 0.89)[[1]] %>% 
  filter(cats__ != "No opinion") %>% 
  mutate(agree = ifelse(cats__ %in% c("Strongly Agree", "Agree"), "Agreeing", "Disagreeing")) %>% 
  ggplot(aes(x = v_380, color = cats__)) +
    geom_point(aes(y = estimate__), size = 2, position = position_dodge(width = 0.2)) +
    geom_errorbar(aes(ymin = lower__, ymax = upper__), width = 0.2, position = position_dodge(width = 0.2)) +
    geom_line(aes(y = estimate__, group = cats__), alpha = 0.7) +
    labs(x = "Frequency of meeting the team", y = "Average probability", color = "Change not communicated") +
    scale_y_continuous(labels = scales::percent) +
    facet_wrap(~agree, ncol = 1) +
    theme_bw()
```

Note that this plot connects the dots (i.e., the average probability per frequency) with lines, which is mathematically incorrect, as the x-axis is ordinal (i.e., categorical) and there is no continuum between two levels.
The connecting lines are solely to visualize the trend better.

Another noteworthy effect is the marginal, i.e., isolated, effect of the way of working on the probability of requirements change not being communicated.

```{r marginal-wow}
conditional_effects(m1, "v_457", categorical = TRUE, prob = 0.89)[[1]] %>% 
  ggplot(aes(x = v_457, color = factor(cats__, levels = cat.likert, ordered = TRUE))) +
    geom_point(aes(y = estimate__), size = 2, position = position_dodge(width = 0.5)) +
    geom_errorbar(aes(ymin = lower__, ymax = upper__), width = 0.3, position = position_dodge(width = 0.5)) +
    labs(x = "Way-of-working", y = "Probability", color = "Change not communicated") +
    scale_y_continuous(labels = scales::percent) +
    theme_bw() +
    theme(legend.position = "bottom")
```

Teams working in an agile way agree perceive requirements not being communicated the least, teams working rather agile the most.
The whiskers strongly overlap, which makes the effect difficult to spot.
Again, a reduced visualization helps detect the effect.

```{r marginal-wow-simplified}
conditional_effects(m1, "v_457", categorical = TRUE, prob = 0.89)[[1]] %>% 
  filter(cats__ != "No opinion") %>% 
  mutate(agree = ifelse(cats__ %in% c("Strongly Agree", "Agree"), "Agreeing", "Disagreeing")) %>% 
  ggplot(aes(x = v_457, color = cats__)) +
    geom_point(aes(y = estimate__), size = 2, position = position_dodge(width = 0.2)) +
    geom_errorbar(aes(ymin = lower__, ymax = upper__), width = 0.2, position = position_dodge(width = 0.2)) +
    geom_line(aes(y = estimate__, group = cats__), alpha = 0.7) +
    labs(x = "Frequency of meeting the team", y = "Average probability", color = "Change not communicated") +
    scale_y_continuous(labels = scales::percent) +
    facet_wrap(~agree, ncol = 1) +
    theme_bw()
```

Thirdly, we can plot the marginal effect of the logarithm of team size (i.e., team size magnitude) on the outcome variable.
For readability, we reduce the credibility intervals from 89% to 50%.

```{r marginal-teamsize}
conditional_effects(m1, "v_3l", categorical = TRUE, prob = 0.5)[[1]] %>% 
  ggplot(aes(x = v_3l, fill = cats__)) +
    geom_line(aes(y = estimate__, color = cats__, group = cats__,)) +
    geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.2) +
    labs(x = "Team size magnitude", y = "Probability", fill = "Response", color = "Response") +
    theme_bw()
```

The larger the magnitude of the team, the more likely a respondent (strongly) agrees with the statement, and the less likely a respondent (strongly) disagrees with the statement of that requirements changes are not communicated.

#### Conditional Effects

Finally, we can visualize the conditional effects, i.e., the interaction between meeting frequency and way-of-working.
These conditional or interaction effects represent an effect moderation, where the effect that the meeting frequency on the probability of requirements changes not being communicated depends on the way-of-working.

```{r conditional-effects-data}
cond <- conditional_effects(m1, effects = "v_380", conditions = data.frame(
                      v_457 = c('Agile', 'Rather agile', 'Hybrid', 'Rather plan-driven', 'Plan-driven')), 
                    categorical = TRUE, prob = 0.89)[[1]] %>% 
  mutate(
    wow = ifelse(v_457 %in% c('Agile', 'Rather agile'), "Agile", 
                      ifelse(v_457 == 'Hybrid', 'Hybrid', 'Plan-driven')),
    cats__ = factor(cats__, 
                    levels = c("Strongly disagree", "Disagree", "Agree", "Strongly Agree"), 
                    ordered = TRUE)
    ) %>% 
    filter(cats__ != "No opinion")
```

The first plot shows the average probabilities of the degree of agreement (color) to the statement that requirements changes are not communicated (y-axis) depending on the meeting frequency (x-axis) for respondents that reported either an `agile` or `rather agile` way of working.

```{r conditional-agile}
cond %>% 
  filter(wow == "Agile") %>% 
  ggplot(aes(x = v_380, color = cats__)) +
    geom_point(aes(y = estimate__), size = 2, position = position_dodge(width = 0.3)) +
    #geom_errorbar(aes(ymin = lower__, ymax = upper__), width = 0.2, position = position_dodge(width = 0.3)) +
    geom_line(aes(y = estimate__, group = cats__), alpha = 0.7, position = position_dodge(width = 0.3)) +
    facet_wrap(~v_457, ncol = 1) +
    labs(x = "Meeting frequency", y = "Probability", color = "Changes not communicated") +
    scale_y_continuous(labels = scales::percent) +
    theme_bw() +
    theme(legend.position = "bottom")
```

These two groups differ rather strongly:

- Respondents working fully agile experience rising probabilities of requirements changes not being communicated when reducing the meeting frequency from *daily* to *weekly*, but it levels out afterwards.
- Respondents working "rather agile" tend to experience problems with changing requirements not being communicated even with daily meetings. Their sour spot is the bi-weekly meeting frequency, though, where the agreement to perceiving requirements changes not being communicated is highest.

Similarly, we can observe the effect for `plan-driven` and `rather plan-driven` projects.

```{r conditional-plan}
cond %>% 
  filter(wow == "Plan-driven") %>% 
  ggplot(aes(x = v_380, y = estimate__, color = cats__)) +
    geom_point(aes(y = estimate__), size = 2, position = position_dodge(width = 0.3)) +
    #geom_errorbar(aes(ymin = lower__, ymax = upper__), width = 0.2, position = position_dodge(width = 0.3)) +
    geom_line(aes(y = estimate__, group = cats__), alpha = 0.7, position = position_dodge(width = 0.3)) +
    facet_wrap(~v_457, ncol = 1) +
    labs(x = "Meeting frequency", y = "Probability", color = "Changes not communicated") +
    scale_y_continuous(labels = scales::percent) +
    theme_bw() +
    theme(legend.position = "bottom")
```

The pattern here is similar for both sub-types of the plan-driven way of working:
Bi-weekly (and in a strict plan-driven way-of-working also: weekly) meetings cause the highest likelihood of perceiving that requirements changes are not communicated.
Both more and less frequent have lower probabilities.

Finally, we can view these same distributions for the `hybrid` way of working.

```{r conditional-hybrid}
cond %>% 
  filter(wow == "Hybrid") %>% 
  ggplot(aes(x = v_380, y = estimate__, color = cats__)) +
    geom_point(size = 2) +
    geom_line(aes(group = cats__), alpha = 0.7) +
    labs(x = "Meeting frequency", y = "Probability", color = "Changes not communicated") +
    scale_y_continuous(labels = scales::percent) +
    theme_bw() +
    theme(legend.position = "bottom")
```

The pattern is again similar, yet shifted.
In a hybrid way-of-working, both bi-weekly and monthly meetings cause the most problems with requirements changes not being communicated.

## Conclusion

The obtained results invite the following conclusions:

- The way-of-working does not explain the non-monotonous effect of meeting frequency on the probability of perceiving that requirements changes are not communicated. Throughout all ways-of-working, there is an intermediate frequency level where the likelihood of requirements changes not being communicated is highest.
- The location of this "sour spot" depends on the way-of-working: for agile teams it is weekly meetings, for rather agile, rather plan-driven, and plan-driven teams it is bi-weekly meetings, and for hybrid teams it is monthly meetings.
- The difference between teams working in an agile versus in a rather agile way is much more pronounced than between any other levels on the way-of-working scale (e.g., plan-driven vs. rather plan-driven).

Note that all of these concluded effects exhibit a certain level of uncertainty, which is visualized with confidence intervals in the results.

[^1]: Siebert, J. (2023). Applications of statistical causal inference in software engineering. Information and Software Technology, 159, 107198.
[^2]: McElreath, R. (2018). Statistical rethinking: A Bayesian course with examples in R and Stan. Chapman and Hall/CRC.
[^3]: Gelman, A., Carlin, J. B., Stern, H. S., & Rubin, D. B. (1995). Bayesian data analysis. Chapman and Hall/CRC.