---
title: "Variance of Techniques for Elicitation"
author: "Julian Frattini"
date: "2026-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse) # R dialect for simpler data wrangling
library(patchwork) # arranging and assembling figures
library(ggdag) # visualizing directed acyclic graphs

library(brms) # training of Bayesian regression models
library(marginaleffects) # evaluation of Bayesian regression models

source("../util/data-loader.R") # preparation of the raw data
```

This notebook contains the analysis of the variance of responses to the options available for the question "Which techniques are employed to elicit requirements in your project?" across demographic factors.
The following options were available to select for this question:

| Code | Name | Text |
|---|---|---|
| `v_36` | Interviews | Interviews |
| `v_37` | Document Analysis | Analysis of existing documents | 
| `v_38` | Risk Analyses | Risk analyses |
| `v_39` | Prototyping | Prototyping | 
| `v_40` | Workshops and Focus Group | Workshops and focus groups | 
| `v_41` | (Requirements) Reuse Databases and Guidelines | (Requirements) Reuse databases and guidelines |
| `v_42` | Design Thinking | Design Thinking |
| `v_43` | External Experts | External experts |
| `v_44` | Observations | Observations |
| `v_47` | Outsourcing | We do not elicit requirements (outsourcing) |
| `v_667` | Usage Data Analysis | Usage data analysis (e.g. event logs) |
| `v_668` | Lean Startup | Lean Startup |
| `v_669` | Experimentation with Users | Experimentation with users |
| `v_461` | I don't know | I don't know |
| `v_454` | None of the Above | None of the above |
| `v_45` | Other | Other |

For analysis, we follow the 3-step framework for statistical causal inference as summarized by Siebert.[^1]

```{r configuration}
# short name of the question
q = "techniques to elicit requirements"
# list of variables representing the options for the question
v <- c("v_36", "v_37", "v_38", "v_39", "v_40", "v_41", "v_42", "v_43",
       "v_44", "v_47", "v_667", "v_668", "v_669", "v_461", "v_454", "v_45")
# names of the options 
labels = c("Interviews", "Document Analysis", "Risk Analyses", "Prototyping", 
           "Workshops and Focus Group", "(Requirements) Reuse Databases and Guidelines",
           "Design Thinking", "External Experts", "Observations", "Outsourcing", 
           "Usage Data Analysis", "Lean Startup", "Experimentation with Users", 
           "I don't know", "None of the Above", "Other")

# minimum likelihood of responses to be considered "common"
level.common <- 0.2
```

## Variables

We consider the following demographic factors:

| Code | Name | Question | Type |
|---|---|---|---|
| `v_23` | country | In which country are you personally located? | categorical |
| `v_3` | size | How many people are involved in your project? | numeric |
| `v_457` | way of working | How would you personally characterise the way your project works? | ordinal |
| `v_1` | sector | Please select the main industrial sector of your project or the application domain of the software you build. | categorical |
| `v_16` | external | Does your project involve external partners? | binary |
| `v_357` | distributed | Is your project distributed? | binary |
| `v_377` | virtualization | Do you tend to work from home, on-site, or both? | ordinal |
| `v_17` | role | What is the main role you occupy in your project? | categorical |

## Data Analysis

We analyze the variance of the response variable following established practices of Bayesian data analysis.[^2]

```{r load-data}
d <- load_data_2024(data_path = "../../data", file = "data_preprocessed.csv") %>% 
  select(
    "lfdn", # respondent identifier
    "v_3", "Country", "v_1", "v_457", "v_16", "v_357", "v_377", "v_17", # demographic factors
    all_of(v) # response variables
    ) %>% 
  drop_na() %>% 
  mutate(v_3 = log(v_3, base = 10))

str(d)
```

For each outcome variable, we train one Bayesian regression model on all demographic variables as predictors.

### Modelling

First, we define a formula regressing the outcome variable of interest on the demographic factors

```{r define-formula}
f <- 
  eval(parse(text = paste0("bf(mvbind(", 
                           paste(v, collapse = ", "),
                           ") ~ 1 + v_3 + Country + v_1 + v_457 + v_16 + v_357 + v_377 + v_17)",
                           "+ set_rescor(FALSE)")))

# get_prior(f, data = d, family = bernoulli())
```

For each coefficient implied by a predictor, we specify a prior based on domain knowledge.

```{r set-priors}
p <- c(
  set_prior("normal(0, 1)", class = "Intercept", resp = gsub("_", "", v)), 
  set_prior("normal(0, 0.2)", class = "b", resp = gsub("_", "", v))
)
```

Since the response variables are all binary (yes/no), we model them using a multivariate Bernoulli distribution utilizing a logit link to map the linear model to the parameter space of $p$.

```{r prior-predictive-check}
m.prior <-
  brm(data = d, family = bernoulli, formula = f, prior = p,
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    seed = 4, sample_prior="only",
    file = paste0("../fits/variance.", v[1], ".prior"), file_refit = "on_change"
  )

brms::pp_check(m.prior, ndraws=100, type = "bars", resp = gsub("_", "", v[1]))
```

The prior predictive check (for one of the response variable options) confirms the plausibility of the selected prior probability distribution.
Given the number of predictors, the weight towards the extreme values is anticipated.
Next, we update the coefficient distributions in light of the collected data using Monte Carlo Markov Chain sampling.

```{r model-training}
m <-
  brm(data = d, family = bernoulli, formula = f, prior = p,
    iter = 2000, warmup = 1000, chains = 4, cores = 4, seed = 4, 
    file = paste0("../fits/variance.", v[1]), file_refit = "on_change"
  )

brms::pp_check(m, ndraws=100, type="bars", resp = gsub("_", "", v[1]))
```

The posterior predictive check (for one of the response variable options) shows that the model predictions ($y_{rep}$) have grown closer to the actually observed data (`y`).
This confirms that the model has trained well.

### Evaluation

We visualize the marginal effect of each demographic factor on every response option to determine, (1) whether an option can be viewed as "common", and (2) whether the levels of the demographic factor affect the responses to this question.
First, we investigate the effect of the sector (`v_1`) on all response options.

```{r marginal-v1, fig.width=12, fig.height=12, unit="cm"}
conditional_effects(m, "v_1", prob = 0.95) %>% 
  purrr::imap_dfr( ~ {
    purrr::map_dfr(.x, identity) %>%
      mutate(response = str_extract(.y, "^[^.]+"))
  }) %>% 
  mutate(
    response = factor(response, levels = gsub("_", "", v), 
                      labels = labels, ordered = TRUE)
  ) %>% 
  ggplot(aes(y = effect1__, color = effect1__)) +
    geom_point(aes(x = estimate__)) +
    geom_errorbar(aes(xmin = lower__, xmax = upper__)) +
    geom_vline(xintercept = level.common, color = "grey", linetype = "dashed") +
    facet_wrap(~response) + 
    scale_x_continuous(labels = scales::percent) +
    labs(x = "Likelihood of selecting the option", y = "Sector") + 
    guides(color = "none") +
    ggtitle(paste0("Sector affecting ", q)) +
    theme_bw()
```

The next figure shows the marginal effect of the project size (`v_3`) on the outcomes.

```{r marginal-v3, fig.width=12, fig.height=12, unit="cm"}
conditional_effects(m, "v_3", prob = 0.95) %>% 
  purrr::imap_dfr( ~ {
    purrr::map_dfr(.x, identity) %>%
      mutate(response = str_extract(.y, "^[^.]+"))
  }) %>% 
  mutate(
    response = factor(response, levels = gsub("_", "", v), 
                      labels = labels, ordered = TRUE)
  ) %>% 
  ggplot(aes(x = effect1__)) +
    geom_line(aes(y = estimate__, color = response)) +
    geom_ribbon(aes(ymin = lower__, ymax = upper__, fill = response), alpha = 0.2) +
    geom_hline(yintercept = level.common, color = "grey", linetype = "dashed") +
    facet_wrap(~response) + 
    labs(x = "Project size magnitude", y = "Likelihood of selecting the option") +
    scale_y_continuous(labels = scales::percent) +
    ggtitle(paste0("Project size magnitude affecting ", q)) +
    guides(color = "none", fill = "none") +
    theme_bw()
```

The following figure shows the marginal effect of the country (`Country`) on the outcomes.

```{r marginal-country, fig.width=12, fig.height=18, unit="cm"}
conditional_effects(m, "Country", prob = 0.95) %>% 
  purrr::imap_dfr( ~ {
    purrr::map_dfr(.x, identity) %>%
      mutate(response = str_extract(.y, "^[^.]+"))
  }) %>% 
  mutate(
    response = factor(response, levels = gsub("_", "", v), 
                      labels = labels, ordered = TRUE)
  ) %>% 
  ggplot(aes(y = effect1__, color = effect1__)) +
    geom_point(aes(x = estimate__)) +
    geom_errorbar(aes(xmin = lower__, xmax = upper__)) +
    geom_vline(xintercept = level.common, color = "grey", linetype = "dashed") +
    facet_wrap(~response) + 
    scale_x_continuous(labels = scales::percent) +
    labs(x = "Likelihood of selecting the option", y = "Country") + 
    guides(color = "none") +
    ggtitle(paste0("Sector affecting ", q)) +
    theme_bw()
```

The next figure shows the variance induced by the way-of-working (`v_457`).

```{r marginal-v457, fig.width=12, fig.height=8, unit="cm"}
conditional_effects(m, "v_457", prob = 0.95) %>% 
  purrr::imap_dfr( ~ {
    purrr::map_dfr(.x, identity) %>%
      mutate(response = str_extract(.y, "^[^.]+"))
  }) %>% 
  mutate(
    response = factor(response, levels = gsub("_", "", v), 
                      labels = labels, ordered = TRUE)
  ) %>% 
  ggplot(aes(y = effect1__, color = effect1__)) +
    geom_point(aes(x = estimate__)) +
    geom_errorbar(aes(xmin = lower__, xmax = upper__)) +
    geom_vline(xintercept = level.common, color = "grey", linetype = "dashed") +
    facet_wrap(~response) + 
    scale_x_continuous(labels = scales::percent) +
    labs(x = "Likelihood of selecting the option", y = "Way-of-working") + 
    guides(color = "none") +
    ggtitle(paste0("Way-of-working affecting ", q)) +
    theme_bw()
```

The next figure shows the variance induced by whether or not the project includes external partners (`v_16`).

```{r marginal-v16, fig.width=12, fig.height=5, unit="cm"}
conditional_effects(m, "v_16", prob = 0.95) %>% 
  purrr::imap_dfr( ~ {
    purrr::map_dfr(.x, identity) %>%
      mutate(response = str_extract(.y, "^[^.]+"))
  }) %>% 
  mutate(
    response = factor(response, levels = gsub("_", "", v), 
                      labels = labels, ordered = TRUE)
  ) %>% 
  ggplot(aes(y = effect1__, color = effect1__)) +
    geom_point(aes(x = estimate__)) +
    geom_errorbar(aes(xmin = lower__, xmax = upper__)) +
    geom_vline(xintercept = level.common, color = "grey", linetype = "dashed") +
    facet_wrap(~response) + 
    scale_x_continuous(labels = scales::percent) +
    labs(x = "Likelihood of selecting the option", y = "External partners") + 
    guides(color = "none") +
    ggtitle(paste0("External partners affecting ", q)) +
    theme_bw()
```

The next figure shows the variance induced by whether or not the project is distributed (`v_357`).

```{r marginal-v357, fig.width=12, fig.height=4, unit="cm"}
conditional_effects(m, "v_357", prob = 0.95) %>% 
  purrr::imap_dfr( ~ {
    purrr::map_dfr(.x, identity) %>%
      mutate(response = str_extract(.y, "^[^.]+"))
  }) %>% 
  mutate(
    response = factor(response, levels = gsub("_", "", v), 
                      labels = labels, ordered = TRUE)
  ) %>% 
  ggplot(aes(y = effect1__, color = effect1__)) +
    geom_point(aes(x = estimate__)) +
    geom_errorbar(aes(xmin = lower__, xmax = upper__)) +
    geom_vline(xintercept = level.common, color = "grey", linetype = "dashed") +
    facet_wrap(~response) + 
    scale_x_continuous(labels = scales::percent) +
    labs(x = "Likelihood of selecting the option", y = "Distributed project") + 
    guides(color = "none") +
    ggtitle(paste0("Distributed project affecting ", q)) +
    theme_bw()
```

The next figure shows the variance induced by the degree of virtualization (`v_377`).

```{r marginal-v377, fig.width=12, fig.height=8, unit="cm"}
conditional_effects(m, "v_377", prob = 0.95) %>% 
  purrr::imap_dfr( ~ {
    purrr::map_dfr(.x, identity) %>%
      mutate(response = str_extract(.y, "^[^.]+"))
  }) %>% 
  mutate(
    response = factor(response, levels = gsub("_", "", v), 
                      labels = labels, ordered = TRUE)
  ) %>% 
  ggplot(aes(y = effect1__, color = effect1__)) +
    geom_point(aes(x = estimate__)) +
    geom_errorbar(aes(xmin = lower__, xmax = upper__)) +
    geom_vline(xintercept = level.common, color = "grey", linetype = "dashed") +
    facet_wrap(~response) + 
    scale_x_continuous(labels = scales::percent) +
    labs(x = "Likelihood of selecting the option", y = "Degree of virtualization") + 
    guides(color = "none") +
    ggtitle(paste0("Degree of virtualization affecting ", q)) +
    theme_bw()
```

The final figure shows the variance induced by the main role of the participant (`v_17`).

```{r marginal-v17, fig.width=12, fig.height=12, unit="cm"}
conditional_effects(m, "v_17", prob = 0.95) %>% 
  purrr::imap_dfr( ~ {
    purrr::map_dfr(.x, identity) %>%
      mutate(response = str_extract(.y, "^[^.]+"))
  }) %>% 
  mutate(
    response = factor(response, levels = gsub("_", "", v), 
                      labels = labels, ordered = TRUE)
  ) %>% 
  ggplot(aes(y = effect1__, color = effect1__)) +
    geom_point(aes(x = estimate__)) +
    geom_errorbar(aes(xmin = lower__, xmax = upper__)) +
    geom_vline(xintercept = level.common, color = "grey", linetype = "dashed") +
    facet_wrap(~response) + 
    scale_x_continuous(labels = scales::percent) +
    labs(x = "Likelihood of selecting the option", y = "Main role in the project") + 
    guides(color = "none") +
    ggtitle(paste0("Main role affecting ", q)) +
    theme_bw()
```

[^1]: Siebert, J. (2023). Applications of statistical causal inference in software engineering. Information and Software Technology, 159, 107198.
[^2]: McElreath, R. (2018). Statistical rethinking: A Bayesian course with examples in R and Stan. Chapman and Hall/CRC.