---
title: 'Hypothesis Hx (Py):'
author: "Julian Frattini"
date: "2025-12-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse) # R dialect for simpler data wrangling
library(patchwork) # arranging and assembling figures
library(ggdag) # visualizing directed acyclic graphs

library(brms) # training of Bayesian regression models
library(marginaleffects) # evaluation of Bayesian regression models

source("../util/data-loader.R") # preparation of the raw data
```

This document investigates hypothesis $H_X^{PY}$ (originating from proposition Y):

> *CAUSE* impacts that *EFFECT*.

The two variables are concepts operationalized by the following self-reported variables:

> *Q1* impacts the *Q2*

For analysis, we follow the 3-step framework for statistical causal inference as summarized by Siebert.[^1]

## Modeling

The causal relationships were formalized in the following directed, acyclic graph (see proposition Y for the full DAG):

```{r specify-dag}
dag <- dagify(
  
)

 ggdag_status(dag, use_labels = "label", text = FALSE) +
   theme_dag() +
   theme(legend.position = "bottom") +
   labs(fill = "Variable type", color = "Variable type")
```

## Identification

Based on the backdoor criterion, we can identify variables to statistically control

```{r identify-adjustment-set}
ggdag_adjustment_set(dag, exposure = "frequency", outcome = "changes")
```

We must adjust for the variable `team.size` to de-confound the effect of `frequency` on `changes`.
The resulting set of relevant variables consists of the following:

| Code | Group | Description | Variable |
|---|---|---|---|
| `frequency` | Communication | How often do members of your team meet, informally by chance meeting or in announced meetings? | `v_380` |
| `changes` | Problems | How much do you agree or disagree with the following statements about your project: Changes in requirements are not communicated | `v_410` |
| `team-size` | Demographics | How many people are involved in your project? | `v_3` |

## Estimation

In the third step, we quantify the strength of the assumed causal effect.
We employ a Bayesian data analysis approach following the guidelines of McElreath[^2] and Gelman et al.[^3]

### Data Preparation

We load the data from the 2024 run of the NaPiRE survey.
The variables relevant to this hypothesis are LIST.
We use the file *data_cleaned_2024.csv*, not the original *data_raw_2024.csv*, because `v_3` is a numeric variable but the original, raw file allowed strings (e.g., "12 people" instead of $12$) in this field.
In the process, we make the following changes:

- Dropping incomplete data: removing all observations containing `NA` values

All manual cleaning operations performed on the data are documented in the data cleaning protocol (to be found in data/runs/napire2024/data-cleaning-protocol.md).

```{r load-data}
d <- load_data(run = "2024", file = "../../data/runs/napire2024/data_cleaned_2024.csv") %>% 
  select(v_3, v_380, v_410) %>%  
  drop_na() %>% 
  mutate(v_3l = log(v_3, base = 10)) %>% 
  filter(v_380 != "I don't know")

str(d)
```

Before starting with the analysis, we can visualize the distribution of the data.
Ordinal variables are visualized as stacked bars.

```{r visualize-data}

```

### Formula

Next, we specify a regression formula based on the identification step.
The outcome variable is ordinal (since it is measured on a 5-point Likert-scale).
We can assume that the ordinal variable is an ordinal operationalization of an unmeasurable latent variables on a continuous scale, i.e., the true agreement to the statement that changes in requirements are not communicated.
Hence, we use the `brms::cumulative` distribution family with a Probit link function.

```{r define-formula}
f1 <- bf(v_410 ~ 1 + v_380 + v_3l)

get_prior(f1, data = d, family = cumulative("probit"))
```

### Priors

We select uninformative priors for all coefficients, i.e., 0-centered normal distributions with wide variance.

```{r set-priors}
p1 <- c(
  set_prior("normal(0, 1)", class = "Intercept"), 
  set_prior("normal(0, 0.5)", class = "b")
)
```

To assess the eligibility of these priors, we perform a prior predictive check.
During this check, we sample from the model using the prior probability distributions for each factor coefficient without updating it in light of the data.

```{r prior-predictive-check}
m1.prior <-
  brm(data = d, family = cumulative("probit"), formula = f1, prior = p1,
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    seed = 4, sample_prior="only",
    file = "../fits/pxhy.prior", file_refit = "on_change"
  )

brms::pp_check(m1.prior, ndraws=100, type = "bars")
```

The sampled prior predictive checks (black whiskers labeled $y_{rep}$) overlap the actually observed data (blue bars labeled $y$), indicating that the selected priors entail the observed reality as realistic.

### Training

With the eligibility of the priors confirmed, we can train the Bayesian regression model, i.e., this time updating the coefficient distributions using Bayes' theorem.

```{r model-training}
m1 <-
  brm(data = d, family = cumulative("probit"), formula = f1, prior = p1,
    iter = 2000, warmup = 1000, chains = 4, cores = 4, seed = 4, 
    file = "../fits/pxhy", file_refit = "on_change"
  )
```

The training does not mention any diverging chains, which indicates solid sampling.
To assess that the model has learned, we perform a posterior predictive check.

```{r posterior-predictive-check}
brms::pp_check(m1, ndraws=100, type="bars")
```

The posterior predictions $y_rep$ still overlap the actually observed data $y$ but the whiskers have grown closer to the data, which indicates that the model has learned.

### Evaluation

Finally, we can evaluate the results to answer the initially posed hypothesis.

#### Posterior Distributions

First, we can investigate the raw posterior distributions obtained from the model training.
These distributions represent the effect that the respective factor has on the outcome variable.
If a distribution is not consistent with 0 (i.e., it does not overlap with 0), then it can be considered statistically significant.

```{r print-posteriors}
summary(m1)
```

These distributions are, however, difficult to interpret, because they are in the parameter space (i.e., on a probit scale), not the outcome space.
To investigate the effects on the outcome space (i.e., in terms of how they affect the likelihood to select a certain category of the Likert-scale), we need to assess the marginal effects.

#### Marginal Effects

To visualize the effect of the predictors on the outcome variable, we plot the marginal effects.
Marginal effects isolate the effect of one predictor while keeping all other predictors at a representative level.


## Conclusion

The obtained results invite the following conclusions:

[^1]: Siebert, J. (2023). Applications of statistical causal inference in software engineering. Information and Software Technology, 159, 107198.
[^2]: McElreath, R. (2018). Statistical rethinking: A Bayesian course with examples in R and Stan. Chapman and Hall/CRC.
[^3]: Gelman, A., Carlin, J. B., Stern, H. S., & Rubin, D. B. (1995). Bayesian data analysis. Chapman and Hall/CRC.